Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach

### 2020-11-25 analysis/paper2/paper2.Rmd
Adds climate dynamics paper

---
twocol: FALSE  #  Use FALSE for submission. 
title: Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
author1: 
  name: Elio Campitelli
  email: elio.campitelli@cima.fcen.uba.ar
author2: 
  name: Leandro B. Díaz
affiliation: |
    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
    CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
    CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
extra_authors: 
  - name: Carolina Vera
    affiliation: |
     Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
     CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
     CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
abstract: |
  Enter the text of your abstract here. This is a sample American Meteorological Society (AMS) \LaTeX\ template.  This document provides authors with instructions on the use of the AMS \LaTeX\ template.  Authors should refer to the file amspaper.tex to review the actual \LaTeX\ code used to create this document. The template.tex file should be modified by authors for their own manuscript.
statement: |
  This is significant becasue I wrote it.
bibliography: ["AsymSAM.bib","packages.bib"]
output: 
  bookdown::pdf_book: 
    base_format: rticles::ams_article
    latex_engine: "xelatex"
  bookdown::word_document2: default
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
 - \usepackage[hidelinks]{hyperref}
---

```{r setup, echo = FALSE,warning=FALSE, message = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.path = "",
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.extra = 42,
  cache.path = "../cache/",
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  out.extra = ""
)

@@

Our method assumes linearity in the asymmetric component of the SAM. That means we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM. @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites. Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures A3 and A4). Therefore, we believe that our method is a reasonable approximation of the phenomenon.

By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON). Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure A5, row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON). In the stratosphere (Figure A5, row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree\ in comparison with the rest of the year. Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA). Based on this, we believe that our initial assumption is not unreasonable throught most of the year with the exception of DJF in the stratosphere.

Finally, we assume that the zonally asymmetric pattern is stationary in time. @silvestri2009 suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes. Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure A6).

@@

The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure A2). This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b]. There is a local peak at 2 years in the periodogram of the tropospheric Symmetric SAM also, although it's not statistically significant. In the troposphere the most significant peak of variability is found in the Asymmetric index at around `r 12*.3` months.

@@

\acknowledgments
NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/

The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.

\datastatement
All data used in this paper is freely available in their respective sources. ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means). NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html). The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt

@@

\bibliography{AsymSAM,packages}

@@

\appendixtitle{Extra figures}

```{r A1, fig.cap = "Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level. Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa. For the 1979 -- 2018 period.", fig.width = 4, fig.height=4, appendix = TRUE}
ccf_dt <- function(..., plot = FALSE) {
  stat <- ccf(..., plot = plot)
  list(lag = stat$lag[, 1, 1], 
       acf = stat$acf[, 1, 1])
}

@@

```{r A2, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5. The shading indicates de 95\\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\\% of the simulated sampels had an amplitude equal or lower). The light line indicates the theoretical expected amplitude from the autorregressive model. For the 1979 -- 2018 period.", appendix = TRUE, fig.height=4, fig.width=7}
spec %>%
  # .[term != "full"] %>% 
  ggplot(aes(1/freq/12, spec)) +
  geom_ribbon(aes(ymin = 0, ymax = `95%`), position = "dodge", alpha = 0.1) +
  geom_line(aes(y = ar_spectrum), alpha = 0.15) +
  geom_line() +
  annotation_logticks(sides = "b") +
  scale_color_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill")) +
  scale_x_log10("Period (years)", sec.axis = sec_axis(~.*12, name = "Period (months)")) +
  scale_y_continuous("Spectrum") +
  facet_grid(lev ~ term,labeller = labeller(lev = lev.lab, term = lab_sam), scales = "free") +
  tag_facets() 
```

@@

```{r A6, fig.cap = "Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018. Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.", fig.width=6, fig.height=5, apendix = TRUE}
set.seed(42)
SAM_prepost <- hgt[lat <= -20] %>% 
  .[lev %in% c(700, 50)] %>%
  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
  .[, year := year(time)] %>% 
  .[, period := factor(ifelse(year(time) <= 1998, "pre", "post"),
                       levels = c("pre", "post"), ordered = TRUE,
                       labels = c("1979-1998", "1999-2018"))] %>%
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
    by = .(period, lev)]

>>>>

---
title: Assessment of zonal symmetric and asymmetric components of the Southern Annular Mode using a novel approach
# subtitle: Do you have a subtitle? If so, write it here
titlerunning:  Assessment of zonal symmetric and asymmetric components of the SAM
# authorrunning: Short form of author list if too long for running head
thanks: | 
    The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.

authors: 
- name: Elio Campitelli
  address: |
    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
      CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
      CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina
  email: elio.campitelli@cima.fcen.uba.ar

- name:  Leandro B. Díaz
  address: |
    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
      CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
      CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina

- name:  Carolina Vera
  address: |
    Universidad de Buenos Aires, Facultad de Ciencias Exactas y Naturales, Departamento de Ciencias de la Atmósfera y los Océanos, Buenos Aires, Argentina
      CONICET – Universidad de Buenos Aires, Centro de Investigaciones del Mar y la Atmósfera (CIMA), Buenos Aires, Argentina
      CNRS – IRD – CONICET – UBA, Instituto Franco‐Argentino para el Estudio del Clima y sus Impactos (UMI 3351 IFAECI), Buenos Aires, Argentina

keywords:
  - Southern Annular Mode
  - general circulation
  - zonally asymmetric circulation
  - El Niño Southern Oscillation

#PACS: 
#- PAC1
#- superPAC

MSC:
- MSC code 1
- MSC code 2

abstract: |
  The text of your abstract.  150 -- 250 words.

bibliography: 
  - AsymSAM.bib
  - packages.bib
bibstyle: spphys
# bibstyle options spbasic(default), spphys, spmpsci
output: 
  bookdown::pdf_book: 
      base_format: rticles::springer_article
      latex_engine: "xelatex"
      # citation_package: "natbib"
header-includes: 
 - \usepackage{gensymb}
 - \usepackage{subfig}
 - \usepackage[inline]{showlabels}
 - \usepackage{chngcntr}
 - \usepackage{natbib}
---

```{r setup, echo = FALSE,warning=FALSE, message = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.path = "",
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.extra = 42,
  cache.path = "../cache/",
  collapse = TRUE,
  comment = "#>",
  dev = "pdf",
  # dpi = 300,
  out.extra = ""
)

@@

Our method assumes linearity in the asymmetric component of the SAM. That means we assume that zonal symmetries associated with positive SAM are almost opposite and of the same magnitude to the ones associated with negative SAM. @fogt2012's composites (their Figure 4) suggest that this might not be entirely valid, although we argue that much of that apparent non-linearity is due to the heterogenous nature of the selected years for constructing the composites. Using our data (from 1979 to 2018), seasonal composites of zonal anomalies of geopotential height for SAM+ (Full SAM index greater than 1 standard deviation) and SAM- (less than minus 1 standard deviation) show relatively high pattern correlations for all seasons and are visually very linear both for the troposphere (represented by the 700 hPa level) and the stratosphere (represented by the 50 hPa level) (Figures \@ref(fig:A3) and \@ref(fig:A4)). Therefore, we believe that our method is a reasonable approximation of the phenomenon.

By computing a single EOF pattern using data for all months we are assuming that the zonal anomalies of the SAM are the same in all seasons -- December to February (DJF), March to May (MAM), June to August (JJA) and September to November (SON). Geopotential zonal anomalies computed by projecting the first EOF of each season independently are very similar to each other in the troposphere (Figure \@ref(fig:A5), row 1) and show pattern correlations between 0.65 (DJF with JJA) and 0.9 (MAM with SON). In the stratosphere (Figure \@ref(fig:A5), row 2), patterns are similar for all trimesters except DJF, when the wave-1 zonal anomalies are rotated 90\degree\ in comparison with the rest of the year. Pattern correlations in the stratosphere are between -0.24 (DJF with SON) and 0.95 (MAM with JJA). Based on this, we believe that our initial assumption is not unreasonable throught most of the year with the exception of DJF in the stratosphere.

Finally, we assume that the zonally asymmetric pattern is stationary in time. @silvestri2009 suggest that this might not be the case between 1958 and 2004 but the period we analyse is much shorter (1979 -- 2018) so it's unlikely that we could observe significant changes. Moreover, zonal asymmetry of the spatial patterns for the two halves of the period (1979 to 1998 and 1999 to 2018) show no systematic change neither in the stratosphere nor in the troposphere (Figure \@ref(fig:A6)).

@@

The stratospheric Symmetric SAM varies strongly with a two-year period, which can be seen by spectral analysis (Figure \@ref(fig:A2)). This might suggests a link between stratospheric SAM variability and the Quasi-Biennial Oscillation [@baldwin2001b]. There is a local peak at 2 years in the periodogram of the tropospheric Symmetric SAM also, although it's not statistically significant. In the troposphere the most significant peak of variability is found in the Asymmetric index at around `r 12*.3` months.

@@

\begin{acknowledgements}
NOAA Global Surface Temperature (NOAAGlobalTemp) data provided by the NOAA/OAR/ESRL PSL, Boulder, Colorado, USA, from their Web site at https://psl.noaa.gov/ 
\end{acknowledgements}

### atastatement
All data used in this paper is freely available in their respective sources. ERA5 data can be obtained via the Copernicus Climate Data Store (https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-pressure-levels-monthly-means). NOAAGlobalTemp and GPCC precipitation data can be obtained through the NOAA Physical Sicences Laboratory website (https://psl.noaa.gov/data/gridded/data.noaaglobaltemp.html and https://psl.noaa.gov/data/gridded/data.gpcc.html). The Oceanic Niño Index is avaiable via NOAA's Climate Precidction Center: https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt

@@

<!-- \bibliography{AsymSAM,packages} -->

@@

\counterwithin{figure}{section}

# Extra figures

```{r A1, fig.cap = "Lag-correlation between Asymmetric SAM and Symmetric SAM index at each level. Negative lags imply Symmetric SAM leading Asymmetric SAM and vice versa. For the 1979 -- 2018 period.", fig.width = 4, fig.height=4, fig.pos="ht"}
ccf_dt <- function(..., plot = FALSE) {
  stat <- ccf(..., plot = plot)
  list(lag = stat$lag[, 1, 1], 
       acf = stat$acf[, 1, 1])
}

@@

```{r A2, fig.cap = "Fourier spectrum of each timeseries computed as Fourier transform smoothed with modified Daniell smoothers with widths 3 and 5. The shading indicates de 95\\% confidence area derived by fitting an autorregressive model and computing the spectrum for 5000 simulated samples from the fitted autoregressive model (95\\% of the simulated sampels had an amplitude equal or lower). The light line indicates the theoretical expected amplitude from the autorregressive model. For the 1979 -- 2018 period.", fig.height=4, fig.width=7}
spec %>%
  # .[term != "full"] %>% 
  ggplot(aes(1/freq/12, spec)) +
  geom_ribbon(aes(ymin = 0, ymax = `95%`), position = "dodge", alpha = 0.1) +
  geom_line(aes(y = ar_spectrum), alpha = 0.15) +
  geom_line() +
  annotation_logticks(sides = "b") +
  scale_color_brewer(NULL, palette = "Dark2", aesthetics = c("color", "fill")) +
  scale_x_log10("Period (years)", sec.axis = sec_axis(~.*12, name = "Period (months)")) +
  scale_y_continuous("Spectrum") +
  facet_grid(lev ~ term,labeller = labeller(lev = lev.lab, term = lab_sam), scales = "free") +
  tag_facets() 
```

@@

```{r A6, fig.cap = "Regression of 50 hPa and 700 hPa geopotential height zonal anomalies (meters) onto the standarised timeseries of the leading EOF computed for the periods 1979 -- 1998 and 1999 -- 2018. Pattern correlation between both fields is 0.86 for the 50 hPa fields and 0.76 for the 700 hPa fields.", fig.width=6, fig.height=5}
set.seed(42)
SAM_prepost <- hgt[lat <= -20] %>% 
  .[lev %in% c(700, 50)] %>%
  .[, hgt := hgt_a*sqrt(cos(lat*pi/180))] %>% 
  .[, year := year(time)] %>% 
  .[, period := factor(ifelse(year(time) <= 1998, "pre", "post"),
                       levels = c("pre", "post"), ordered = TRUE,
                       labels = c("1979-1998", "1999-2018"))] %>%
  .[, .(eof = list(EOF(hgt ~ time | lon + lat, n = 1, data = .SD))), 
    by = .(period, lev)]
